---
title: "Practical 3 Answers"
subtitle: "Working with Data: Exploratory Data Analysis"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

<!--
Before you start:

- This practical uses the `penguins` dataset from the `palmerpenguins`
  package — there's no .csv to download this week, just an install
  (see the README for instructions).
- See the README.md in this repository for package installation instructions
  and Git/GitHub tips.
- Render this document with the "Render" button in RStudio, or by running
  quarto::quarto_render("practical3-answers.qmd") in the R console.
-->

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

# Load packages
library(ggplot2)
library(dplyr)
library(naniar)
library(palmerpenguins)
library(GGally)

penguins
```

## Palmer Penguins

### Question 1

Use the `naniar` package to inspect missingness. Are there any variables where missingness is high and this should be monitored?

_You can use `naniar::gg_miss_upset(...)` or other functions you've come across but you should carefully interpret the output, and review the raw data to confirm.^[There may be some warnings that you can safely ignore.]_

```{r}
#| label: inspect-missingness
#| 
naniar::miss_var_summary(penguins)
naniar::vis_miss(penguins)
naniar::gg_miss_upset(penguins)
```

> Missingness is low overall (0.7% of all cells). There are two rows where most of the data is missing. A further 7 rows have sex missing.

> If you’re interested in why some data is missing, check the penguins_raw dataset, in particular the Comments column.


### Question 2

Summarise the data variables in the dataset. What variables exist? What are their ranges, means etc? Hint: you can start by using `glimpse()` and `summary()`. You may also look at the whole dataset using `View()` in RStudio.

```{r}
#| label: summarise-variables

glimpse(penguins)
summary(penguins)
penguins |> count(species)
penguins |> count(island)
penguins |> count(sex)
penguins |> count(year)
```

> penguins has 344 rows and 8 columns, including:

> Three categorical variables
  - species: Adelie/Chinstrap/Gentoo
  - island: Biscoe/Dream/Torgersen
  - sex: female/male

> Four continuous measurements
  - bill_length_mm,bill_depth_mm
  - flipper_length_mm
  - body_mass_g
  - year: 2007/2008/2009 (effectively categorical here even though it’s stored as a number).

> Note that species is imbalanced (152 Adelie, 124 Gentoo, 68 Chinstrap).


## Basic Plotting

### Question 3

Using `geom_histogram()`, plot the distribution of `flipper_length_mm`. What shape does the distribution appear to have? Use the starter code below and change the theme with `theme_bw()`.

```{r}
#| label: histogram-flipper-length

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram() +
  theme_bw()
```

> The distribution is not unimodal (e.g., like a bell curve). It looks to be bimodal, or even trimodal, with a cluster of shorter flippers (~180-195mm) and a separate cluster of longer flippers (~210-230mm), and a less distinct middle. This is a hint that flipper length differs systematically by some other variable in the data rather than representing one homogeneous population.


### Question 4

The default bin width is rarely ideal. Experiment with the `binwidth` argument in `geom_histogram(...)`. How does your choice of bin width change your interpretation of the distribution?

```{r}
#| label: histogram-binwidth

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 10) +
  ggtitle("binwidth = 10") +
  theme_bw()

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 2) +
  ggtitle("binwidth = 2") +
  theme_bw()

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 0.5) +
  ggtitle("binwidth = 0.5") +
  theme_bw()
```

> Choice of bin width can hide or reveal structure in the data:
- Too wide a bin (e.g. binwidth = 10, meaning collect in 10mm bins) smooths right over the multiple clusters and can make the data look unimodal.
- A moderate bin width (e.g. binwidth = 2) is what reveals the multimodal shape clearly.
- Too narrow a bin (e.g. binwidth = 0.5) startsshowing gaps and noise from individual measurement values.

### Question 5

Using `geom_bar()`, visualise the count of penguins belonging to each species. Which species is most common in this dataset?

```{r}
#| label: bar-species-count

ggplot(penguins, aes(x = species)) +
  geom_bar() +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  theme_bw()

```
> Adelie is the most common species (152 penguins), followed by Gentoo (124) and then Chinstrap 
>(68, under half as many as Adelie).

### Question 6

Modify your bar chart to map the `fill` aesthetic to `island`. What does the resulting stacked bar tell you about where each species was sampled? Try `position = "dodge"` in `geom_bar()`, which version is appropriate and when?

```{r}
#| label: bar-species-island

# Stacked (default position = "stack")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  scale_fill_discrete("Island") +ggtitle("Stacked version") +theme_bw()

# Dodged
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "dodge") +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  scale_fill_discrete("Island") +
  ggtitle("Dodged version") +
  theme_bw()

# Dodged version 2
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = position_dodge(preserve = "single")) +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  scale_fill_discrete("Island") +
  ggtitle("Dodged 2.0") +
  theme_bw()

```

> Both charts show that species and island are not independent: Gentoo were only sampled on Biscoe, Chinstrap only on Dream, while Adelie were sampled on all three islands. This means island is confounded with species.

>  Stacked: Better for showing the total count and how it is divided by island.
>  Dodged: Better for directly comparing island counts across species.

> Here, dodged is clearer for comparing groups, using position_dodge(preserve = "single") to avoid misleading bar widths.


### Question 7

Use `geom_boxplot()` to compare the distribution of `body_mass_g` across the three species. Map `species` to the x-axis and `body_mass_g` to the y-axis.

```{r}
#| label: boxplot-mass-species

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  scale_x_discrete("Species") +
  scale_y_continuous("Body mass (g)") +
  theme_bw()
```

> Gentoo penguins are clearly heavier than Adelie and Chinstrap, whose distributions overlap quite a bit with each other.

### Question 8

Map the `fill` aesthetic in your previous boxplot to `sex`. What happens to the number of boxes, and what new comparison does this enable?

```{r}
#| label: boxplot-mass-species-sex

ggplot(penguins, aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  scale_x_discrete("Species") +
  scale_y_continuous("Body mass (g)") +
  scale_fill_discrete("Sex") +
  theme_bw()
```

> The number of boxes triples: instead of one box per species (3 boxes), we now get one box per species-sex combination.

> We can see that within every species, males are likely to be heavier than females.

> This pattern was invisible in the previous plot, where sex differences were averaged away inside each species’ single box.


### Question 9

Remove the missing values in the previous plot by filtering. Is this good practice? When should missing values be removed?

```{r}
#| label: boxplot-remove-missing

penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  scale_x_discrete("Species") +
  scale_y_continuous("Body mass (g)") +
  scale_fill_discrete("Sex") +
  theme_bw()

```

> Is it good practice, and when should you remove missing values? It depends!
It is bad practice to remove NAs silently, without checking why values are missing or whether the missingness is related to some other variable of interest (e.g. if sex were harder to determine for smaller or younger birds, dropping those rows could bias comparisons, not just reduced the sample size).
A safer strategy is to keep the missing values visible where possible, and only explicitly filter when you have a specific reason.
Here, as we’re doing exploratory data analysis, including the NAs makes sense as we want to note their missingness. But in a polished report we may exclude NA rows and reference their removal in the text. It always depends on what you would like the plot to demonstrate to the reader. Ask yourself: “what story am I trying to tell?”


## Advanced Plotting

### Question 10

Recreate the above plot using `geom_point()`. Describe the overall relationship between these two variables and its dependence on species.

```{r}
#| label: scatter-mass-species

penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, colour = sex)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  scale_x_discrete("Species") +
  scale_y_continuous("Body mass (g)") +
  scale_colour_discrete("Sex") +
  theme_bw()
```

> Using “jitter” moves the points horizontally a random amount, ensuring no data points are hidden. The same overall story as the boxplot holds: average body mass for Adelie and Chinstrap and approximately equal, whilst Gentoo is higher. Then within every species, the male points sit visibly higher than the female points, with some overlap.

> The scatter also makes individual outliers and the density of points within each group more visible than the boxplot’s five-number summary does, at the cost of being a bit more cluttered to read at a glance. The boxplot also has the advantage of being able to read off the summary statistics (e.g. median and some quartiles).


### Question 11

Change the previous plot so that each species is plotted separately using `facet_wrap()`. When would using facets be more appropriate than colours to distinguish categories? Try using the `scales = "free"` argument for `facet_wrap()` — when would this be helpful or inappropriate?

```{r}
#| label: facet-species

penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = sex, y = body_mass_g, colour = sex)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  facet_wrap(~species) +
  scale_x_discrete("Sex") +
  scale_y_continuous("Body mass (g)") +
  scale_colour_discrete("Sex") +
  theme_bw()
```

> It’s much easier to see the female vs male spread with the facet plots.

> Facets become more appropriate when

>   the number of groups grows (colour palettes get hard to distinguish past ~5-6 categories)
    groups don’t overlap much so splitting them out loses little context
    when overlapping points/ distributions from several groups would otherwise obscure each other (the case now)


```{r}
# Same plot as before, but letting each facet's y-axis scale independently
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = sex, y = body_mass_g, colour = sex)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  facet_wrap(~species, scales = "free") +
  scale_x_discrete("Sex") +
  scale_y_continuous("Body mass (g)") +
  scale_colour_discrete("Sex") +
  theme_bw()
```
> Using scales = "free" can be useful when the scales vary widely between facets, or the comparison is only required within each facet. It should always be used carefully and noted in the text describing the graph. Here it is highly likely to misrepresent the data! A quick glance makes it look as if Gentoo have similar body mass to Adelie and Chinstrap.


### Question 12

Use the `ggscatmat()` function from `GGally` to plot a scatter plot matrix of pairwise relations between continuous variables.

```{r}
#| label: scatmat-continuous

GGally::ggscatmat(penguins, columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"))
```

> The function ggscatmat yields, in one plot, every pairwise scatterplot between the four continuous variables (below the diagonal), each variable’s univariate density (on the diagonal), and each pairwise correlation coefficient (above the diagonal). It is a fast way to screen for pairwise relationships, without writing out six separate geom_point() calls by hand.

### Question 13

Add colouring by species to the `ggscatmat()` plot. From the plot, identify which two variables are likely to best distinguish each species. Are there any anomalies in the data?

```{r}
#| label: scatmat-species-colour

GGally::ggscatmat(
  penguins,
  columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"),
  color = "species"
)
```

> Bill length and bill depth together appear to separate the three species most cleanly1.

> Gentoo have shallow-but-long bills, Adelie have short-but-deep bills, and Chinstrap sit apart from Adelie mainly on bill length despite similar depth. The three species form three fairly distinct, only lightly-overlapping clusters on this pair.

> In terms of anomalies; a small number of points sit outside their species’ otherwise tight cluster. These could be genuine anomalies (measurement error or individual variation) or related to a sub-populations like juveniles. During EDA you should flag and investigate such anomalies rather than ignoring or dropping them.

